---
title: 'Case Study Report: Bellabeat Wellness Insights'
author: "Aakif Altaf"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    theme: united
    toc: true
    toc_float: true
    highlight: tango
  word_document:
    toc: true
---



---

# Introduction

This case study explores **smart device usage data** from Fitbit to uncover patterns in activity, sleep, and wellness behavior.  
Our goal is to generate actionable insights that Bellabeat, a wellness technology company, can use to improve its **marketing strategies and product offerings**.  

The analysis follows a structured data analysis approach:  
- **Ask**: What patterns exist in activity, sleep, and calories burned?  
- **Prepare**: Clean and merge Fitbit datasets.  
- **Process**: Explore and visualize relationships between variables.  
- **Analyze**: Identify correlations and trends.  
- **Share**: Present findings with clear plots and narrative.  
- **Act**: Recommend strategies for Bellabeat’s products and campaigns.  


---

# Libraries & Data Import

```
Imported and explored multiple CSV files using readr and dplyr

Cleaned and merged datasets using distinct(), drop_na(), and inner_join()

Verified data integrity and structure (wide vs. long format)

Assessed credibility using the ROCCC framework (Reliable, Original, Comprehensive, Current, Cited)
```

```{r setup, message=FALSE, warning=FALSE,echo=FALSE}
library(tidyverse)
library(lubridate)
library(skimr)
library(janitor)

# Import data
daily_activity <- read.csv("C:/Users/Aaqib/Downloads/test_data/dailyActivity_merged.csv")
sleep_day <- read.csv("C:/Users/Aaqib/Downloads/test_data/sleepDay_merged.csv")
weight_log<- read.csv("C:/Users/Aaqib/Downloads/test_data/weightLoginfo_merged.csv")

# Standardize column names
daily_activity <- clean_names(daily_activity)
sleep_day <- clean_names(sleep_day)
weight_log <- clean_names(weight_log)

# Remove duplicates
sleep_day <- sleep_day %>% distinct()

# Convert date columns
daily_activity <- daily_activity %>% mutate(activity_date = mdy(activity_date))
sleep_day <- sleep_day %>% mutate(sleep_day = mdy_hms(sleep_day))

# Quick data summary
skim_without_charts(daily_activity)
skim_without_charts(sleep_day)

```
 **I standardized and cleaned the datasets. For analysis, I focus mainly on daily activity and sleep records, merging them where possible.**
 
---

# Analysis & Insights

**Steps vs Calories**


**Question: Do more steps translate to more calories burned?**

```{r, echo=FALSE}

ggplot(daily_activity, aes(x= total_steps, y = calories)) +
  geom_point(color = "blue", alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Steps vs Calories Burned", x = "Total Steps", y = "Calories Burned")


```

**Insight:**

Users who took more steps consistently burned more calories. The relationship is strong and linear, confirming that physical activity drives calorie expenditure.

**Business Implication:**

Bellabeat can emphasize step challenges and activity goals in its marketing to highlight calorie-burning benefits.

---
# Sleep vs Activity

**Question: How does sleep duration affect activity levels?**
```{r, echo=FALSE}
activity_sleep <- daily_activity %>%
  inner_join(sleep_day, by = c("id" = "id", "activity_date" = "sleep_day"))

ggplot(activity_sleep, aes(x= total_minutes_asleep, y = total_steps)) +
  geom_point(color = "darkred", alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  labs(title = "Sleep vs Activity", x = "Minutes Asleep", y = "Steps")

```


**Insight:**

Longer sleep often correlates with lower daily activity. This suggests a trade-off between rest and physical movement.

**Business Implication:**

Bellabeat can encourage balanced routines—using notifications to promote consistent sleep while nudging users toward more daily movement.


# Weekly Activity Patterns

**Question: Are there patterns across weekdays?**
```{r,echo=FALSE}
daily_activity %>%
  mutate(weekday = wday(activity_date, label = TRUE)) %>%
  group_by(weekday) %>%
  summarise(avg_steps = mean(total_steps)) %>%
  ggplot(aes(x = weekday, y = avg_steps, fill = weekday)) +
  geom_col(show.legend = FALSE) +
  labs(title = "Average Steps by Day of Week", x = "Day", y = "Average Steps")

```

**Insight:**

Users are more active on weekdays compared to weekends. Step counts drop notably on Saturdays and Sundays.

**Business Implication:**

Bellabeat could introduce weekend challenges or reminders to maintain activity
levels outside of workdays.


# Conclusions & Recommendations

**This analysis highlights opportunities for Bellabeat to improve user engagement and promote healthier lifestyles:**

**Encourage weekend activity challenges to balance weekday vs weekend movement.**

**Use personalized notifications to promote sleep consistency and healthy routines.**

**Highlight calorie–step link in marketing to promote fitness through walking/activity tracking.**

**Promote balanced wellness by showing users how to manage both sleep and activity effectively.**

**By leveraging these data-driven insights, Bellabeat can enhance user wellness outcomes while strengthening its brand as a holistic smart health companion.**




# Thank You

Thank you for reviewing this case study.
I appreciate your time and interest in my analysis.
If you would like to discuss this project further or explore potential collaborations, feel free to connect with me on my Email: kmrtech99@gmail.com
